### 2013-01-10 paper/chemokine_main.tex
Final submitted to PLoS CB

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals constraints on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.  T cell navigation to this region is undirected but migration is made possible by local cytokine and chemokine production from infected epithelial cells.  We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using an agent-based model (CyCells).

Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used to induce chemokine production \textit{in vitro}.  Of the chemokines tested, secretion of CXCL10 (IP-10) and CCL5 (RANTES) were stimulated by infection, with IP-10 dominating the migration due to higher concentrations.  A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.  We observe that T cell sensitivity to chemokine is high enough to maximize their efficiency.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.

The spatial nature of the model reveals unique challenges to T cell recruitment not visible in standard differential equation models.  Infected cells can become isolated in expanding plaques, impeding infection control.  In the model, T cells executed an efficient search with sensitivity to either chemokine.  A key limitation imposed on T cells was illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci as revealed in cinematics. This spatial model is consistent with efficient T cell recruitment to infected lung.  Without an apparent means to induce increased efficiency, vaccination remains the key to shortening the interval between initial infection and T cell activation.

@@

The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that control viral replication of all but the most virulent strains.  Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.   The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.   Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.  The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.   On approximately day 5 after infection activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.

@@

The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.   For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediated T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediated the opposite effect \cite{DeLemos2005}.  In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of CXCL10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.   In the herpesvirus model CXCL10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.  In the parainfluenza virus model CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.  In the influenza A model initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.   The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency but viral clearance was not altered \cite{Fadel2008}.  The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.

Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, T cells have not be shown to climb chemokine gradients.   Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.   Finally the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.

A number of mathematical models have constructed the entire adaptive response and key components \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.  Focused math models can examine each event in detail and narrow the range of possible solutions.  We sought to examine the CD8 T cell diaspora and recruitment to infected lung in detail.  Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail.   Activated T cells searching for and homing into infected tissue do so in a spatially complex environment.  We therefore utilized the strategy of agent-based modeling (ABM) to represent the physical environment of the searching T cell.  Thus T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.

We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.  Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only compare simulated outcome among three influenza strains.  To achieve this focused goal on the efficacy of the chemotactic signals alone, we use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.  Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Models}

@@

We estimate chemokine production rates by adapting the delay differential equation model described in \cite{Mitchell2011} by adding one new equation to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_1[t-\tau_2] \\
\dot{C} &= r I[t-\tau_3] - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells, creating a ball of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.

@@

\section*{Materials}

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF was ignored as its effects are outside the scope of this paper.

@@

The model simulates a chemokine gradient surrounding an infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).  T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.

Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (single runs were used due to computational limitations, model variance is discussed in S2.1.).  We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  Refer to the videos in Figures S3-S5 to see a representative simulation of seasonal influenza.

In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, secreting cells were not eliminated in either H1N1 simulations (Fig. S1).  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.  The video in Figure S6 illustrates this event.

@@

Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.  Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.  Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.  The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.  All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.  Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.  Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.  Thus, our model is not intended to predict clearance of virus from the lung.  Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.

The use of an ABM has certain advantages over a spatially homogeneous ODE model.  An ODE model assumes that any virus particle is capable of infecting any healthy cell.  Figure \ref{fig:cycells} shows that this is clearly not the case.  In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  This is limiting as can be seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.

Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).  Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlier.  First, because T cells find infected cells by climbing a gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.   Second, T cell clustering persists after all virus-secreting cells have been eliminated.  The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.  Finally, we can see that infected cells are more disperse as infection size grows.  Because T cells are clustered they cannot cover the increasing plaque effectively.  These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.   The behavior of searching T cells described in this paper 
can enhance future global immune response models.

@@

% - removed - Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?  The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (S2.2). Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity. T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance. We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency. Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.

@@

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{SystemChart}
\end{center}
\caption{{\bf Model of T cell search.}  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through 14 vascular bifurcations of the bronchial network.  Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.  In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
\label{fig:systemchart}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{ModelChart}
\end{center}
\caption{{\bf Visual representation of the model.}  Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.  Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.  Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.  Solid arrows represent a cell state transition from one behavior to another.  Dashed arrows display the mechanism used to induce a transition.  Dotted arrows indicate the production of new virus.}
\label{fig:modelchart}
\end{figure}

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.  T cell navigation to infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.  We focused on the efficiency of local chemokines to induce migration of activated CD8 T cells using a spatially explicit agent-based model (ABM) without constructing a model containing all elements of the immune response.

To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.  A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.

The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.  Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.  A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci. This spatially explicit model is consistent with efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.

@@

The adaptive immune response induced during acute influenza A infection is a complex combination of defense mechanisms.  Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.   The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.   Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.  The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.   On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.

@@

The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.   For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediates T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediates the opposite effect \cite{DeLemos2005}.  In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of IP-10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.   In the herpesvirus model, IP-10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.  In the parainfluenza virus model, CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.  In the influenza A model, initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.   The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency, but viral clearance was not altered \cite{Fadel2008}.  The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.

Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.   Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.   Finally the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.

There are many mathematical models of the entire adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.  Mathematical and computational models can examine each event in detail and narrow the range of possible behaviors.  Here we examine the CD8 T cell diaspora and recruitment to infected lung in detail.  Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.   Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment.  We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.  In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.

We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.  Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.  We use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.  Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

@@

We estimate chemokine production rates by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} by adding one new equation to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_1[t-\tau_2] \\
\dot{C} &= r I[t-\tau_3] - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen levels of bifurcations with the origin at the bronchial lymph node complex. Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node.  When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in the capillary bed at random locations.

@@

\subsection*{Materials}

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF was ignored as its effects are outside the scope of this paper.

@@

The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).  T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.

Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (model variance is discussed in Text S1 2.1.).  We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  Refer to the Videos S1-S3 to see a representative simulation of seasonal influenza.

In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, secreting cells were not eliminated in either H1N1 simulations (Fig. S1).  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.  Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.  Video S4 illustrates this event.

@@

The model makes a number of simplifications to and deletions of elements in the innate and adaptive immune response, allowing us to build a tractable model where data was available.  Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.  Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.  The contribution of IgM antibody clearing free virus is represented as a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.  All of these rates may in fact be time-variable as indicated by data-fitting models \cite{Wu2011}.  Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.  Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.  Thus, our model is not intended to predict clearance of virus from the lung, but to examine the response  features that permit CD8 T cells to sense and contact infected target cells.

The use of an ABM has certain advantages over a spatially homogeneous ODE model.  An ODE model assumes that any virus particle is capable of infecting any healthy cell.  Figure \ref{fig:cycells} shows that this is clearly not the case for viral adhesion and entry.  In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  This is limiting as can be seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.

Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.  Because CD8 T cells find infected cells by presumably climbing a chemokine gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.   We also see that T cell clustering persists after all virus-secreting cells have been eliminated.  The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.  Finally, we can see that infected cells are more disperse as infection size grows.  Because T cells are clustered they cannot cover the increasing plaque effectively.  These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM (\textit{in vivo} pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation).   The behavior of searching T cells described in this paper can enhance future global immune response models.

@@

% - removed - Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?  The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (Text S1 2.2). Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity. T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance. We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency. Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.

@@

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{Figure_1}
\end{center}
\caption{{\bf Model of T cell search.}  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through 14 vascular bifurcations of the bronchial network.  Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.  In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
\label{fig:systemchart}
\end{figure}

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{Figure_2}
\end{center}
\caption{{\bf Visual representation of the model.}  Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.  Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.  Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.  Solid arrows represent a cell state transition from one behavior to another.  Dashed arrows display the mechanism used to induce a transition.  Dotted arrows indicate the production of new virus.}
\label{fig:modelchart}
\end{figure}
